Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-30 pnas/Towards.tex
no message

The signal to noise ratio (SNR) of the projections in the micrographs is a function of two dominating factors. On the one hand, the SNR is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the SNR is a function of the molecule size. The smaller the molecules, the fewer detected electrons carry information about them. \TODO{This isn't quite accurate: it's about the density of particles in the image; but we want to make a point about size; how do we reconcile this?}

@@

As a proof of concept, we study
\TODO{This sentence beginning makes it sound like what follows is our main point; should the main point be the 3D recovery, even if low-res?}
a toy model for which it is easier to convey the mathematical principles. 
In this toy model, an unknown image appears multiple times at unknown locations in each of several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is challenging in particular when the SNR is low enough that particle picking (identifying the locations of each image in each micrograph) cannot be done reliably. 
This problem is interesting of its own as it appears in other scientific applications, including spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

@@

To support this claim, we consider a strictly simpler problem: suppose an oracle identifies for us intervals of length $L$ in the micrograph that either contain a full signal occurrence (plus noise), or contain just noise. Our task is to determine which intervals contain signal, that is, to determine whether the corresponding entries of $s$ are 0 or 1.
The oracle further provides the signal $x$, the probability $q$ that one of the intervals contains signal, and the noise variance $\sigma^2$. \TODO{Nico: I don't see how the next sentence is relevant for the task of estimating $s$---Since we have the full generative model, we can always generate more data as needed; therefore it is enough to consider the hardness of deciding whether a single interval contains signal or not.---I would just write this: Since the intervals are independent of each other, it is sufficient to consider a single interval.}

The decision problem for a single interval can be abstracted as follows: We have two known vectors $\theta_0 = x$ and $\theta_1 = 0$ in $\mathbb{R}^L$. There is a random variable $\eta$ taking values 0 or 1 with probabilities $q$ and $1-q$, respectively. We observe a random vector $X \in \mathbb{R}^L$ (an extract of the micrograph) with the following distribution:
% $X | (\eta  = i) \sim N(\theta_i,\sigma^2)$; in other words,
if $\eta = 0$, then $X \sim N(\theta_0,\sigma^2)$ (that is, $X$ follows a normal distribution around $\theta_0$ with variance $\sigma^2$), and if $\eta = 1$, then $X \sim N(\theta_1,\sigma^2)$.
%If $\theta_1=0$, then the latter corresponds to pure noise interval.

We observe $X$, and our task is to estimate $\eta$. How reliably can this be done? If $q \geq 1/2$, the constant estimator $\hat{\eta} = 0$ is correct with probability $q$; likewise for $\hat{\eta} = 1$ if $q \leq 1/2$, correct with probability $1-q$. The question is, can we do better than this? We prove that, as $\sigma \to \infty$, the answer is no:
\begin{proposition} \label{prop:two_gauss}
	For any deterministic estimator $\hat{\eta }$ of $\eta$,
	\begin{align*}
	\lim_{\sigma \to \infty} \P[ \hat{\eta} = \eta  ] & \leq \max(q, 1-q);
	\end{align*}
	that is: as the SNR deteriorates, the probability of success is no better than random chance.
\end{proposition}
\noindent The result is proved in Appendix~\ref{sec:proof_two_gauss}.

@@

Under the separation condition on $s$,  for shifts in $0, \ldots, L-1$, the autocorrelation functions of $y$ depend only on the corresponding autocorrelations of $x$, the density $\gamma$ and the noise level $\sigma$. In particular, for the first-order autocorrelation (that is, the mean), since the noise is independent of the signal, the mathematical expectation of $a_y^1$ is easily seen to be:
\begin{align*}
\E\{ a_y^1 \} & = \gamma a_x^1.
\end{align*}
(The randomness is over the Gaussian noise $\varepsilon$; $s$ may be deterministic: under the conditions, the quantity under consideration depends on $s$ only through $\gamma$.)

We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density.) In that regime, the law of large numbers can be used to show the following statement:
\begin{align} \label{eq:mean_micrograph}
\lim_{N\to\infty} a_y^1 & = \gamma a_{x}^1.
\end{align}
Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)

We have a similar observation for the second-order autocorrelation: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$), and never with another occurrence. As a result,
\begin{align}
	\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell]
	\label{eq:ac2_micrograph}
\end{align}
for $\ell = 0, \ldots, L-1$, where $\delta$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.

@@

\begin{figure*}
    \centering
    \includegraphics[angle=270, width=0.3\textwidth]{TRPV1_rec.eps}
    \caption{Reconstruction of the TRPV1 molecule from its clean first 3 autocorrelations. The reconstruction is shown in gray while the ground truth is shown in yellow.}
    \label{fig:TRPV1_recon}
\end{figure*}

>>>>

The signal to noise ratio (SNR) of the projections in the micrographs is a function of two dominating factors. On the one hand, the SNR is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the SNR is a function of the molecule size. The smaller the molecules, the fewer detected electrons carry information about them. \TODO{This isn't quite accurate. Amit: do you agree with the last sentence? maybe we should say it is a function of the contrast?}

@@

As a proof of concept, we first study a toy model \TODO{Amit: do you want to keep ``toy model'' or replace with ``mathematical toy model'', ``simplified model'' etc.} for which it is easier to convey the mathematical principles. 
In this toy model, an unknown image appears multiple times at unknown locations in each of several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is challenging in particular when the SNR is low enough that particle picking (identifying the locations of each image in each micrograph) cannot be done reliably. 
This problem is interesting of its own as it appears in other scientific applications, including spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

@@

To support this claim, we consider a strictly simpler problem: suppose an oracle identifies for us one interval of length $L$ in the micrograph that either contains a full signal occurrence (plus noise), or contains just noise. Our task is to determine which one it is, that is, to determine whether the corresponding entry of $s$ is 0 or 1.
The oracle further provides the signal $x$, the probability $q$ that  the interval contains signal, and the noise variance $\sigma^2$. %\TODO{Nico: I don't see how the next sentence is relevant for the task of estimating $s$---Since we have the full generative model, we can always generate more data as needed; therefore it is enough to consider the hardness of deciding whether a single interval contains signal or not.---I would just write this: Since the intervals are independent of each other, it is sufficient to consider a single interval.}

This decision problem can be abstracted as follows: We have two known vectors $\theta_0 = x$ and $\theta_1 = 0$ in $\mathbb{R}^L$. There is a random variable $\eta$ taking values 0 or 1 with probabilities $q$ and $1-q$, respectively. We observe a random vector $X \in \mathbb{R}^L$ (an extract of the micrograph) with the following distribution:
% $X | (\eta  = i) \sim N(\theta_i,\sigma^2)$; in other words,
if $\eta = 0$, then $X \sim \mathcal N(\theta_0,\sigma^2)$ %(that is, $X$ follows a normal distribution around $\theta_0$ with variance $\sigma^2$),
 and if $\eta = 1$, then $X \sim \mathcal N(\theta_1,\sigma^2)$. %\TODO{Tamir: to remove parenthesis}
%If $\theta_1=0$, then the latter corresponds to pure noise interval.

We observe $X$, and our task is to estimate $\eta$. How reliably can this be done? If $q \geq 1/2$, the constant estimator $\hat{\eta} = 0$ is correct with probability $q$; likewise, if $q \leq 1/2$, the constant estimator $\hat{\eta} = 1$ is correct with probability $1-q$. The question is, can we do better than this?  We prove that, as $\sigma \to \infty$, the answer is no:
\begin{proposition} \label{prop:two_gauss}
	For any deterministic estimator $\hat{\eta }$ of $\eta$,
	\begin{align*}
	\lim_{\sigma \to \infty} \P[ \hat{\eta} = \eta  ] & \leq \max(q, 1-q);
	\end{align*}
	that is: as the SNR deteriorates, the probability of success is no better than random chance.
\end{proposition}
\noindent The result is proved in Appendix~\ref{sec:proof_two_gauss}.

@@

For shifts in $0, \ldots, L-1$, the autocorrelation functions of $y$ depend on the corresponding autocorrelations of $x$, the noise level $\sigma$ and the support signal $s$. Importantly, under the separation condition~\eqref{eq:spacing}, the dependency on $s$ is only through the density $\gamma$.
%In particular, for the first-order autocorrelation (that is, the mean), since the noise is independent of the signal, the mathematical expectation of $a_y^1$ is easily seen to be:
%\begin{align*}
%\E\{ a_y^1 \} & = \gamma a_x^1.
%\end{align*}
%(The randomness is over the Gaussian noise $\varepsilon$; $s$ may be deterministic: under the conditions, the quantity under consideration depends on $s$ only through $\gamma$.)
We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density). In that regime, the law of large numbers can be used to show the following statement:
\begin{align} \label{eq:mean_micrograph}
\lim_{N\to\infty} a_y^1 & = \gamma a_{x}^1.
\end{align}
The randomness is over the Gaussian noise $\varepsilon$; $s$ may be deterministic.
Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)

We have a similar observation for the second-order autocorrelation: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$)\TODO{Tamir: remove parenthesis}, and never with another occurrence. As a result,
\begin{align}
	\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell],
	\label{eq:ac2_micrograph}
\end{align}
for $\ell = 0, \ldots, L-1$, where $\delta$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.

@@

\begin{figure}
    \centering
    \includegraphics[angle=270, width=0.4\textwidth]{TRPV1_rec.eps}
    \caption{Reconstruction of the TRPV1 molecule from its clean first 3 autocorrelations. The reconstruction is shown in gray while the ground truth is shown in yellow.}
    \label{fig:TRPV1_recon}
\end{figure}
